\documentclass[12pt]{extarticle}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{natbib}
\usepackage{gensymb}
\bibliographystyle{apa}

\title{Parental Effects on Development and Behavior in Threespine Stickleback \textit{Gasterosteus aculeatus} of Lake M\'yvatn, Iceland}
\author{Spencer Edwards, Roll No.}
\date{April 2021}

\begin{document}

\maketitle

\section*{Introduction}
Development and early life history are two of the most important time periods for an organism. Consequentally, the traits gained by an organism from its parents during this time are key to its success as an individual. It is not suprising then, that parental effects, both maternal and paternal, have a major impact on the an organism from before it is born \citep{charmantier_garant_kruuk_2014, Danchin2011, Badyaev2009}. Parental effects, defined as a change in an Offspring's phenotype due to the genotype and/or environment of one or more of its parents, were for a long time considered a nuicance, a deviation from pure heritable traits. Now however we have seen a researgence of interest in these effects \citep{charmantier_garant_kruuk_2014}. While much research has been done concerning the impact of parental effects on development \citep{Tigreros2021} and behavior of invertabrates (particularly beetles), comparitivley less is known about parental effects on vertabrates. Work on teleosts has revealed strong links between parental effects on the co-evolution of behavior and cellular function \citep{Yoshizawa2012}. The connection between parental affects influencing behavior and gene expression has serious implications for evolutionary ecology, because rapid environmental could act as a mechanism for rapid evolution via parental programming of offspring phenotypes, effectively providing their offspring with a ``jumpstart'' of sorts \citep{Danchin2011, Donelson2018}. 
From the standpoint of quantitative genetics, the phenotype (\textit{P}) is equal to the influence of genes (\textit{G}) and environment (\textit{E}) such that $$P = G \times E $$

However, genetic effects from parents can be further broken down into inheritence from genes directly ($I_G$) and inheritence from the parental environment ($I_E$) such that $$P = (I_G \times I_E) \times E$$

Here in Iceland, work on various populations of \textit{Gasterosteus aculeatus} has focused on phenotypic diversity, including phenotypic plasticity \citep{Kristjansson2002, Millet2013}. However, the extent to which parental effects shape the morphology and behavior of lake M\'yvatn stickleback has yet to be investigated. Previous work on different stickleback populations have discovered a range of parental effects, both maternal and paternal. \citet{Bell2018} investigated the heritability of parental behavior (Specifically, fanning of eggs) by male stickleback and found strong heritability of the trait. Furthermore, they concluded that a strong amount of genetic variation could lead the evolvability of the fanning trait. Offspring of male stickleback that expirience predation risk have shown to grow smaller and spend less time in the open \citep{Bell2016, Stein2014} Female stickleback have also been shown to pass on phenotypic information to their eggs. An RNAseq study analyzing female sitckleback found that eggs from mothers exposed to predators had faster development times, as well as major epigenetic changes and alterations to non-coding genes during development \citep{Mommer2014, Bell2016}.  \\


Major questions I want to address:
\begin{itemize}
 \item To what extent are the phenotypes of lake M\'yvatn stickleback 
shaped by maternal and paternal (parental) effects?
 \item Specifically, how much variation in stickleback phenotypes are driven by genetic inheritence vs environmental inheritence?
 \item Do M\'yvatn stickleback 
\end{itemize}


\section*{Methods}
I will use a half-sibiling common garden expiriment to assess TGP in M\'yvatn stickleback. Stickleback will be collected from seperate basins of the lake, the North, or cold shore, and the South, or warm shore. Sticklebacks from both shores will then be crossed using standard methods laid out by Schluter. Half-sibling split clutch designs have been used previously to examine the effects of indirect fitness in parasite resistance, as well as transgenerational plasticity in response to temperature in marine stickleback \citep{Barber2001,Ramler2014}. One of the major problems when considering historical work on parental effects is disentangling genetic and environmental effects \citep{Donelson2018}. Using a Linear Mixed Model (LMM) approach to decompile the effects should be effective.
15 males and 15 breeding females will be collected from the shore of the hot basin (hereon: hot shore) and a nearby cool region called Grimstaðir. Sperm from each male will be split and used to fertilize two females, and eggs from each female will be fertilized by two males. Each family will be split between two tanks to account for tank effects on individuals. Fish will be
\\

Proposed Guides: Dr. Bjarni Kristj\'ansson, Department of Aquaculture and Fish Biology, H\'olar University College.
\bibliography{mainbib}

\end{document}

